---
title: "Diffing two pipeline runs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diffing two pipeline runs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The question this package exists to answer is "what does this change to
OMOP-ES do to its output?". None of the code below is evaluated when the site
is built, since it all needs an OMOP-ES checkout and access to the source
databases.

## The short version

```{r}
library(omopesutils)

omop_es_diff_viewer_local_git(
  omop_es_path = "~/omop_es",
  left_branch = "main",
  right_branch = "my-feature-branch",
  links_patient_id_column = "my_patient_id_column"
)
```

That runs the pipeline twice, loads both extracts, and opens the diff viewer.
The rest of this article takes the same thing apart, which is what you want
when a step fails, or when the two extracts already exist.

## Running the pipeline

`omop_es_run()` runs a pipeline in a fresh subprocess. The subprocess matters:
the OMOP-ES scripts `source()` each other and assign into the global
environment, so running two versions in one session would let the first
contaminate the second.

```{r}
omop_es_run(
  omop_es_path = "~/omop_es",
  settings_id = "CUH_EPIC_small_cohort",
  cohort_limit = 5000,
  custom_dir = "~/omop_es/extract/diff/left"
)
```

`omop_es_run_git_sha()` wraps this with the git handling: it stashes any
uncommitted changes, checks out the branch you name, fast-forwards it to its
upstream, runs the pipeline, and pops the stash. It deliberately refuses to
merge anything that is not a fast-forward, rather than trying to resolve
divergence.

```{r}
omop_es_run_git_sha(
  branch = "my-feature-branch",
  omop_es_path = "~/omop_es",
  custom_dir = "~/omop_es/extract/diff/right"
)
```

Note that this leaves the checkout on the branch it last ran.

## Loading both extracts into one database

A comparison needs both extracts queryable at once, which means one duckdb
connection with two pairs of schemas. Nothing is copied --- each table becomes
a view over the parquet or CSV files on disk.

```{r}
db <- DBI::dbConnect(duckdb::duckdb())

duckdb_register_omop_es_output(
  db,
  extract_path = "~/omop_es/extract/diff/left/CUH_EPIC_small_cohort_2026-02-01",
  omop_es_path = "~/omop_es",
  schema_public = "dbo",
  schema_private = "priv"
)

duckdb_register_omop_es_output(
  db,
  extract_path = "~/omop_es/extract/diff/right/CUH_EPIC_small_cohort_2026-02-01",
  omop_es_path = "~/omop_es",
  schema_public = "dbo2",
  schema_private = "priv2"
)
```

The public/private split is not cosmetic. OMOP-ES writes the `_links` tables,
which carry source-system identifiers, separately from the OMOP tables
themselves, and `duckdb_register_omop_es_output()` keeps that separation by
putting them in different schemas.

`dbo`/`priv` and `dbo2`/`priv2` are the defaults the diff viewer expects, so
using those names saves passing them again below.

## Looking at the difference

```{r}
omop_es_diff_viewer(
  db,
  links_patient_id_column = "my_patient_id_column"
)
```

The viewer has three panels, in decreasing order of how quickly they answer
"did anything change?":

- **Table Row Counts** --- one row per OMOP table, with the count on each side
  and the change between them. A table that appears on only one side shows
  `NA`, which is how you spot a table that has started or stopped being
  populated.
- **Plugin Row Counts** --- the same, broken down by the plugin that produced
  each row, so a change can be attributed to a particular mapper rather than
  just to a table.
- **Details** --- a row-level diff of a single table for selected patients,
  rendered by [daff](https://github.com/edwindj/daff).

Surrogate keys are not expected to be stable between pipeline runs, so the
Details panel drops them before comparing; otherwise every row would look
changed. This is `drop_omop_foreign_keys = TRUE` in
`omop_es_tbl_with_links()`.

## Working with the tables directly

The viewers are built on `omop_es_tbl_with_links()`, which is worth knowing
about on its own: it returns a lazy `tbl` for an OMOP table with the relevant
`_links` tables already joined on, columns prefixed so they cannot collide.
It is internal, hence the `:::` below.

```{r}
omopesutils:::omop_es_tbl_with_links(
  db,
  "condition_occurrence",
  schema_public = "dbo",
  schema_private = "priv"
) |>
  dplyr::count(links__plugin_provenance)
```

Which `_links` tables are relevant is worked out from the OMOP CDM
specification shipped with the package, rather than hard-coded: the table's
own `_links` table, plus those of every table it references by a
non-vocabulary foreign key. That is why a `condition_occurrence` query can be
filtered by a source-system patient identifier that only exists on `person`.

## Row counts without the app

The counts behind the first two panels are available directly, which is handy
in a script or a regression check.

```{r}
omopesutils:::omop_diff_tables_row_count(
  db,
  schema_left = "dbo",
  schema_right = "dbo2"
)

omopesutils:::omop_diff_plugins_row_count(db)
```
